Subglacial and subaerial fluvial sediment transport capacity respond differently to water discharge variations

### 2024-03-21 Delaneyetal2024.tex
small changes

\section*{Plain Language Summary} % 200 words
Sediment export from glaciers impacts downstream ecosystems and communities and has changed in response to climate warming and glacier retreat.
Sediment export depends in part on the sediment transport capacity, and the amount of sediment that could be transported for the hydraulic conditions with abundant sediment availability.
This capacity largely depends on the width of the channel and the water's velocity.
In subaerial channels, like rivers, water velocity, and channel width both respond rapidly to water discharge.
However, subglacial channels are pressurized by the ice above, so water flows through a subglacial pipe that evolves in size.
This pressurized flow means that channel size evolves slowly compared to variations in water discharge, therefore these variations are largely accommodated by changes in water velocity as opposed to channel size.
Models of these processes show that sediment transport capacity is highly variable in subglacial channels, and a wide range of water discharges can result in comparable sediment transport capacities.
In subaerial channels, by contrast, sediment transport capacity is less variable, and variations depend solely on water discharge, not the channel size.
The channel's different responses to water discharge may impact the interpretation of sediment transport records from glacierized catchments, complicating the link between glacier melt and sediment discharge.

@@

  \begin{figure}[h]
  \centering
    \includegraphics[width=0.8\linewidth]{Fig1.pdf}
    \caption{Sketch for the different short time-scale responses of subglacial and subaerial channels to increased water discharge.
      Water velocity magnitudes in the subglacial, $v_g$, and subaerial, $v_f$, channels are shown by arrow length.
      $S_g$ and $S_f$ represent the wetted area in subglacial and subaerial channels, respectively.
      The subglacial channel width $w_g$ remain unchanged, while the subaerial channel width $w_f$ evolves. Shear stress $\tau$ is responsible for the mobilization of sediment.
    }
    \label{fig:cartoon}
  \end{figure}

@@

Water discharge data is used from Alpine and ice sheet settings collected downstream of the glacier, assuming that no water storage occurred in the proglacial area.
The Alpine site (\alpine{}) is  Fieschergletscher in the Swiss Alps ($46^\circ\,29'\,07''$ N, $8^\circ\,08'\,3''$ E).
The water discharge data used here was collected at a $1$\unit{min} interval from May 24, 2014 to October 10, 2014
see \citeA{felix2022} and \citeA{felix2021} for more details.

@@

\section{Methods}
\label{sect:meth}
The models described below (Sections ~\ref{sect:sub_mode}~and~\ref{sect:fluv}) represent relationships amongst water discharge, water velocity, and channel geometry in both subaerial and subglacial channels (Table \ref{table:vpm}).
Both models use the measured discharge to calculate water velocity, shear stress, and width-integrated shear stress, upon which sediment transport both suspended load and bedload depend \cite<Figure \ref{fig:cartoon}; >{shields1936}.
The evaluation of these variables omits the selection of a sediment transport relationship and a grain-size parameter \cite<e.g.>{shields1936,meyer1948} needed to calculate sediment transport capacity from shear stress.

@@

The models above are applied to proglacial discharge records from the Fieschergletscher (scenario \alpine{}) and the Leverett Glacier (scenario \icesheet{}).
The subglacial and subaerial models represent the subglacial flow of water exiting a glacier and transitioning to subaerial flow.
Outputs of the models represent generalizable sediment transport characteristics from these hydrographs, rather than actual hydraulic conditions.
To generalize these scenarios, \alpine{}  is exemplified by relatively thin ice thickness ($h_{ice}$: $225$\,\unit{m}), low water discharge ($\sim\,10$\,\unit{m}$^3$\,\unit{s}$^{-1}$) and high diurnal variability in water discharge (Figure~\ref{fig:model_outs}, a).
\icesheet{}  is exemplified by thick ice  ($h_{ice}$: $700$\,\unit{m}), high water discharge ($\sim\,300$\,\unit{m}$^3$\,\unit{s}$^{-1}$)  and low diurnal variability in water discharge (Figure~\ref{fig:model_outs}, e).
%
However, both scenarios share a discharge with a similar, relative seasonal amplitude.

For each glacier, the models are first applied to a reference test case for each glacier that assumes  a subglacial channel with $\beta=\frac{\pi}{6}$, i.e. a segment of a quarter-circle, (Table~\ref{table:vpm}).
Both friction factors $f_r$ and $f_f$ are tuned so that reasonable water velocities occur for both \alpine{} and \icesheet{} \cite<$\sim\,1.6\,$\unit{m}$^3$\,\unit{s}$^{-1}$>{werder2010b,chandler2013}.
The initial condition of cross-sectional area $S_g$ is evaluated by applying a model spinup with the maximum water discharge for the first $3$ days of the study until the change in channel area is negligible. 
The subaerial channels of both scenarios have a bed slope of $\nabla z_c = 0.05$ (Table~\ref{table:vpm}).

The results below present water velocity ($v_g$, $v_f$), shear stress ($\tau_g$, $\tau_f$), and width-integrated shear stress ($w_g\tau_g$, $w_f\tau_f$) from the subglacial and subaerial models.
For brevity, these values together are referred to as ``model outputs'' in the text.
Outputs  are compared to water discharge from the glaciers using Spearman rank correlation, which varies with the ordering of values.
Rank correlation reduces the impact of the non-linear response of sediment transport capacity to the hydrological forcing, thus making our interpretation independent of the exact details of the sediment transport formula.

@@

To establish the role of water discharge variability in driving variations in sediment transport capacity, the water discharge records are aggregated across a select period ranging from $15$\,\unit{min} up to $15$\,\unit{days}.
Aggregation periods longer than this  are omitted as they might be conflated with seasonal variations in water discharge.
Here, the water discharge input at a given time is the average over the aggregation period before that time.  
The routine runs until $100$ different parameter combinations are accepted for each water discharge aggregation period.

@@

\subsection{Role of water discharge in sediment transport capacity}
\begin{figure}[h]
  \centering
  \includegraphics[width=0.9\linewidth]{Fig2.pdf}
  \caption{Model outputs from simulations using the hydrographs in panels a and e for the scenarios \alpine{} (a-d) and \icesheet{} (e-h).
    Gray (black) lines in a and e represent the subglacial channel size (water discharge).
    Blue lines represent outputs from the subglacial channel, while orange lines represent the subaerial channel.
    Data are shown at $15$\unit{min} intervals, i.e. not aggregated.
    Arrows denote where variables peak in the subglacial channel before the subaerial one.
    Insets in e-h show the  peak melt event \icesheet{} denoted by the shaded area in panel e.
    % Note that the x and y axes are different for \alpine{} and \icesheet{}.
  }
  \label{fig:model_outs}
\end{figure}

@@

\begin{figure}[h]
  \centering
  \includegraphics[width=0.9\linewidth]{Fig3.png}
  \caption{
    Relationship between water discharge and normalized velocity, shear stress, and width-integrated shear stress for \alpine{} (a-c) and \icesheet{} (d-f).
    Variables on the x-axis have been normalized to mean values.
    $R_g$ and $R_f$ show the Spearman rank correlations for the subglacial and the subaerial outputs, respectively.
    Plots are shown with data and outputs at $15$\unit{min} intervals.
  }
  \label{fig:Qw_vari}
\end{figure}

@@

\begin{figure}[h]
  \centering
    \includegraphics[width=0.9\linewidth]{Fig4.png}
    \caption{Variability (standard deviation) of model outputs for different time aggregations across the range of subglacial and subaerial channel shapes and friction factors.
      Shaded areas denote the range of results from the accepted $100$ parameter combinations tested (Section~\ref{sect:imp}).
      Solid lines denote  mean values.
      %Top row shows results from the \alpine{} scenario, while bottom row show outputs from the \icesheet{} scenario
      Time intervals shown with markers denote $15$\,\unit{min}, $1$\,\unit{hr}, $12$\,\unit{hr}, $1$\,\unit{d}, $5$\,\unit{d}, $10$\,\unit{d}, and $15$\,\unit{d} aggregation periods.
       }
    \label{fig:multi_run}
  \end{figure}

The second numerical experiment aims to evaluate the role of channel shape and friction factors on the findings above and to examine the effect of variability in water discharge.
Across the range of channel shapes and friction factors  examined (Section\,\ref{sect:imp}), variability in water velocity and shear stress remains higher in the subglacial system compared to the subaerial one in both \alpine{} and \icesheet{}.
This suggests that there is likely greater intermittency in sediment transport in subglacial systems compared to subaerial ones, especially in supply-limited systems where sediment could be exhausted or mobilized over diurnal discharge cycles.
In contrast, in \alpine{}, the variability in width-integrated shear stress is comparable between the subaerial and subglacial systems, due to the highly variable water discharge, largely due to diurnal variations (Figure~\ref{fig:multi_run}).
While the seasonal variability is comparable, substantial variations in sediment discharge likely persist between the two channel types from the different timing of peak discharge events.
However, in \icesheet{}, with lower water discharge fluctuations, the variability in width-integrated shear stress remains consistently higher in the subglacial system across the time aggregation periods.

Evaluating the different variability in sediment transport characteristics across timescales can help inform the sediment transport processes omitted from applying daily or weekly values of water discharge to understand sediment transport.
Results from  both \alpine{} and \icesheet{} scenarios show that the variability in subglacial  decreases with aggregations longer than approximately 1--5 days (Figure~\ref{fig:multi_run}).
Yet, substantial differences in the variability between the model outputs for subglacial and subaerial cases persist in periods of up to $15$ days.
Note again that the timing of peak sediment transport conditions still occurs across these periods, and they are are thus not captured in aggregating water discharge over longer periods (Figure~\ref{fig:model_outs}).

\section{Discussion}
\subsection{Hydrology and evolving channel shape in driving sediment transport variability}
\label{sect:dis_qsc}

@@

Sediment transport capacity in subglacial channels responds more rapidly to changes in water discharge compared to their subaerial counterparts \cite{alley1997}.
In subglacial R\"othlisberger channels in steady state with water discharge, the width-integrated shear stress response to water discharge is $ w_g\,\tau_{g}\, \propto\,  Q^{\sim\frac{4}{5}}$ (Table\,\ref{tab:eqs1}).
However, subglacial channels rarely operate in a steady state \cite<e.g.>{gimbert2016}.
For sufficiently short periods, channel width and wetted area are fixed in the subglacial system (akin to pipe flow) so that the response of width-integrated shear stress is $w_g\, \tau_{g}\, \propto\,  Q^{2}$ (Table\,\ref{tab:eqs1}).
\ref{sect:scaling} provides a complete description of sediment transport capacity in different drainage conditions.
Note that in pipe flow conditions, maximum sediment transport would coincide with peak water discharge and a co-varying relationship with water discharge would occur, unlike the case here with evolving channel size (Figure \ref{fig:model_outs} and \ref{fig:Qw_vari}). 
However, similar adjustments in channel shape occur in subaerial channels \cite{phillips2016}, albeit at  timescales longer than hours to days in subglacial channels.

 Substantial differences emerge in model outputs between the different hydrological forcings of the \alpine{} and \icesheet{} cases.
Indeed, the more independent relationship between model outputs and water discharge in \alpine{} results from the
relatively larger discharge variations on daily to weekly timescales compared to \icesheet{}.
This occurs as the evolution of the R-channel's size is small compared to the variations in water discharge.
Such water discharge variability in alpine catchments may cause shear stress to approach the water discharge squared \cite< \ref{sect:scaling}; c.f.>{alley1997}.
With respect to \icesheet{}, a closer approximation of the hydraulics could be represented by pipe flow here, where the channel size remains fixed.
Conversely, the reduced variability in water discharge in \icesheet{} may result in the R-channel's size being closer to equilibrium concerning the variations in water discharge.
In this case, the exponent on water discharge for shear stress is likely less than $2$, but greater than $\frac{4}{5}$ that occurs when the R-channel's size evolves in equilibrium with water discharge (see \ref{sect:scaling}).
As a result, a stronger correlation persists between subglacial model outputs and water discharge in \icesheet{} compared to \alpine{} (Figure~\ref{fig:Qw_vari}).

Very large quantities of sediment are known to be evacuated during glacier lake drainages or intense precipitation events \cite<e.g.>{cowan1988,delaney2018}.
In these cases, massive amounts of sediment can be evacuated from the subglacial environment, due to increased sediment access, but also the large increases in sediment transport capacity.
If water input before these events have been minimal, the channel size would be small and sediment transport capacity would very rapidly increase at the onset of the event (see \ref{sect:scaling}).

@@

The stronger link between sediment transport capacity and water discharge in subaerial channels may mean that variations in sediment transport records can be more easily attributed to hydro-climatic conditions than in subglacial systems.
Over time scales of weeks or longer, the glacier's sediment transport capacity is impacted by the ice thickness, controlling the channel closure rate, and the glacier's surface slope, in addition to water discharge \cite<Figure~\ref{fig:multi_run}, Section~\ref{sect:sub_mode}; >{rothlisberger1972,shreve1972,gimbert2016,stevens2022}.
In a supply-limited state, then glaciers' sediment discharge record also responds to  sediment availability and bedrock erosion from  water pressure variations and sliding  \cite<e.g.>{iverson2012,herman2015,delaney2023}.
This multitude of processes lies in contrast to processes occurring in most subaerial channels, where transport capacity typically responds to sediment size, channel shape, water discharge, and hydraulic gradient.
The hydraulic gradient and channel shape can remain stable over years to millennia \cite<Section~\ref{sect:fluv}; e.g.>{tucker1997,wickert2019}.

@@

Furthermore, the greater variations in subglacial sediment transport capacity in both catchment types compared to subaerial ones could cause a supply-limited regime to persist at many glaciers.
This regime may persist as the sediment's threshold of motion can be reached more frequently and across a range of water discharges, compared to subaerial systems (Figure~\ref{fig:Qw_vari}).
Sediment exhaustion occurring through increased variability may also explain the stronger dependence of sediment discharge from the Greenland Ice Sheet on basal shear stress, a proxy for bedrock erosion, rather than glacier melt \cite{overeem2017}.

\subsection{Implications for parameterizing  sediment transport in glacierized catchments}

Numerical models of subglacial sediment transport generally apply classical parameterizations of subglacial hydrology to evaluating sediment transport conditions \cite<e.g>{hewitt2019,beaud2018}.
In these R-channel models, viable water pressure regimes can be achieved by changing the shape and friction factor of the channels, enabling assumptions of semi-circular channels in most applications \cite{werder2010b}.
However, the role of channel width in the findings here suggests that more thorough parameterization of channel shape and channel width is needed in assessing subglacial sediment transport capacity as compared to evaluating water pressure (Section\,\ref{sect:sub_mode}, \ref{sect:scaling}).

In applying these models to different climate scenarios, water discharge measurements at the sub-hour timescale remain unavailable in many catchments and daily values might be used, for instance.
The variability in subglacial channels approaches that of subaerial ones  over $1-5$ days (Figure~\ref{fig:multi_run}), possibly resulting in parity in variability between subaerial and subglacial channels over these periods.
This suggests that aggregating water discharge over daily to weekly periods could yield similar variations between subglacial and subaerial systems, although generally steeper hydraulic gradients of glaciers may cause elevated sediment transport capacity there (Figure\,\ref{fig:model_outs}). 
At the same time, water discharge over periods longer than days may omit key processes, such as the maximum sediment transport capacity occurring with increased water discharge, as opposed to the maximum quantity (Figure~\ref{fig:model_outs}).

@@

The sediment transport capacity of a channel is driven by its width and the shear stress exerted by the flowing water, which, in turn, is proportional to the square of the flow velocity.
Subaerial channels alter both their channel width and water velocity immediately in response to changing water discharge.
In contrast, pressurized subglacial channels largely accommodate rapidly changing water discharge by altering water velocity.
Their size responds over several days to changes in discharge.
Thus the response of sediment transport capacity is much more sensitive to changes in water discharge in subglacial channels compared to subaerial ones.

Implications of the presented findings include:
\begin{enumerate}
\item the changing relationship between water discharge and sediment transport capacity underneath glaciers creates larger variability in sediment transport capacity compared to river systems;
  \vspace{.1cm}

\item the timing of peak sediment transport capacity below glaciers generally coincides with large rates of increase in water discharge, as opposed to the maximum water discharge in subaerial channels.
  However, representing subglacial sediment transport capacity with either open-channel flow or pipe flow results in coincident variations; 
  \vspace{.1cm}

\item the degree of difference in sediment transport capacity variations between the two channel types depends on the water discharge variability in the hydrograph, which varies with catchment characteristics and water discharge evolution;
  \vspace{.1cm}

  \item the uneven response of subglacial sediment transport capacity to water discharge variations shows a different response to periods of elevated water discharge in subglacial channels compared to subaerial ones.
\end{enumerate}

>>>>

\section*{Plain Language Summary} % 200 words
Sediment export from glaciers impacts downstream ecosystems and communities and has changed in response to climate warming.
Sediment export depends partly on the sediment transport capacity, and the amount of sediment that could be transported for the hydraulic conditions with abundant sediment availability.
This capacity largely depends on the width of the channel and the water's velocity.
In subaerial channels, like rivers, water velocity, and channel width both respond rapidly to water discharge.
However, subglacial channels are pressurized by the ice above, so water flows through a subglacial pipe that evolves in size.
This pressurized flow means that channel size evolves slowly compared to variations in water discharge, therefore these variations are largely accommodated by changes in water velocity as opposed to channel size.
Models of these processes show that sediment transport capacity is highly variable in subglacial channels, and a wide range of water discharges can result in comparable sediment transport capacities.
In subaerial channels, by contrast, sediment transport capacity is less variable, and variations depend solely on water discharge, not the channel size.
The channel's different responses to water discharge may impact the interpretation of sediment transport records from glacierized catchments, complicating the link between glacier melt and sediment discharge.

@@

  \begin{figure}[h]
  \centering
    \includegraphics[width=0.8\linewidth]{Fig1.pdf}
    \caption{Sketch for the different short time-scale responses of subglacial and subaerial channels to increased water discharge.
      Arrow length denotes Water velocity magnitudes in the subglacial, $v_g$, and subaerial, $v_f$, channels.
      $S_g$ ($S_f$) represent the wetted area in subglacial (subaerial) channels.
      The subglacial channel width $w_g$ remain unchanged, while the subaerial channel width $w_f$ evolves.
      Shear stress $\tau$ is responsible for the mobilization of sediment.
    }
    \label{fig:cartoon}
  \end{figure}

@@

Water discharge data is used from Alpine and ice sheet settings collected downstream of the glacier, assuming no water storage in the proglacial area.
The Alpine site (\alpine{}) is  Fieschergletscher in the Swiss Alps ($46^\circ\,29'\,07''$ N, $8^\circ\,08'\,3''$ E).
The water discharge data used here was collected at a $1$\unit{min} interval from May 24, 2014 to October 10, 2014
see \citeA{felix2022} and \citeA{felix2021} for more details.

@@

\section{Methods}
\label{sect:meth}
The models described below (Sections ~\ref{sect:sub_mode}~and~\ref{sect:fluv}) represent relationships amongst water discharge, water velocity, and channel geometry in both subaerial and subglacial channels (Table \ref{table:vpm}).
Both models use the measured discharge to calculate water velocity, shear stress, and width-integrated shear stress, upon which both suspended sediment and bedload transport depend \cite<Figure \ref{fig:cartoon}; >{shields1936}.
The evaluation of these variables omits the selection of a sediment transport relationship and a grain-size parameter \cite<e.g.>{shields1936,meyer1948} needed to calculate sediment transport capacity from shear stress.

@@

The models above are applied to proglacial discharge records from the Fieschergletscher (scenario \alpine{}) and the Leverett Glacier (scenario \icesheet{}).
The subglacial and subaerial models represent the subglacial flow of water exiting a glacier and transitioning to subaerial flow.
Outputs of the models represent generalizable sediment transport characteristics from these hydrographs, rather than actual hydraulic conditions.
To generalize these scenarios, \alpine{}  is exemplified by relatively thin ice thickness ($h_{ice}$: $225$\,\unit{m}), low water discharge ($\sim\,10$\,\unit{m}$^3$\,\unit{s}$^{-1}$) and high diurnal variability in water discharge (Figure~\ref{fig:model_outs}, a).
\icesheet{}  is exemplified by thick ice  ($h_{ice}$: $700$\,\unit{m}), high water discharge ($\sim\,300$\,\unit{m}$^3$\,\unit{s}$^{-1}$)  and low diurnal variability in water discharge (Figure~\ref{fig:model_outs}, e).
%However, both scenarios share a discharge with a similar, relative seasonal amplitude.

For each glacier, the models are first applied to a reference test case for each glacier that assumes  a subglacial channel with $\beta=\frac{\pi}{6}$%, i.e. a segment of a quarter-circle,
(Table~\ref{table:vpm}).
Both friction factors $f_r$ and $f_f$ are tuned so that reasonable water velocities occur for both \alpine{} and \icesheet{} \cite<$\sim\,1.6\,$\unit{m}$^3$\,\unit{s}$^{-1}$>{werder2010b,chandler2013}.
The initial condition of cross-sectional area $S_g$ is evaluated by applying a model spinup with the maximum water discharge for the first $3$ days of the study until the change in channel area is negligible. 
The subaerial channels of both scenarios have a bed slope of $\nabla z_c = 0.05$ (Table~\ref{table:vpm}).

The results below present water velocity ($v_g$, $v_f$), shear stress ($\tau_g$, $\tau_f$), and width-integrated shear stress ($w_g\tau_g$, $w_f\tau_f$) from the subglacial and subaerial models.
For brevity, these values together are referred to as ``model outputs'' in the text.
Outputs  are compared to water discharge from the glaciers using Spearman rank correlation, which varies with the ordering of values.
Rank correlation reduces the impact of the non-linear response of sediment transport capacity to the hydrological forcing.

@@

To establish the role of water discharge variability in driving variations in sediment transport capacity, the water discharge records are aggregated across a select period ranging from $15$\,\unit{min} up to $15$\,\unit{days}.
Here, the water discharge input at a given time is the average over the aggregation period before that time.  
The routine runs until $100$ different parameter combinations are accepted for each water discharge aggregation period.

@@

\subsection{Role of water discharge in sediment transport capacity}
\begin{figure}[h]
  \centering
  \includegraphics[width=0.9\linewidth]{Fig2.pdf}
  \caption{Model outputs from simulations using the hydrographs in panels a and e for the scenarios \alpine{} (a-d) and \icesheet{} (e-h).
    Gray (black) lines in a and e represent the subglacial channel size (water discharge).
    Blue (orange) lines represent outputs from the subglacial (subaerial) channel.
    Data are shown at $15$\unit{min} intervals.
    Arrows denote where variables peak in the subglacial channel before the subaerial one.
    Insets in f-h show the  peak melt event \icesheet{} denoted by the shaded area in panel e.
    % Note that the x and y axes are different for \alpine{} and \icesheet{}.
  }
  \label{fig:model_outs}
\end{figure}

@@

\begin{figure}[h]
  \centering
  \includegraphics[width=0.9\linewidth]{Fig3.png}
  \caption{
    Relationship between water discharge and normalized velocity, shear stress, and width-integrated shear stress for \alpine{} (a-c) and \icesheet{} (d-f).
    Variables on the x-axis have been normalized to mean values.
    $R_g$ ($R_f$) show the Spearman rank correlations for the subglacial (subaerial) outputs.
    Plots are shown with discharge and outputs at $15$\unit{min} intervals.
  }
  \label{fig:Qw_vari}
\end{figure}

@@

\begin{figure}[h]
  \centering
    \includegraphics[width=0.9\linewidth]{Fig4.png}
    \caption{Standard deviation of model outputs for different time aggregations with different subglacial and subaerial channel shapes and friction factors.
      Shaded areas denote the range of results from the accepted $100$ parameter combinations (Section~\ref{sect:imp}).
      Solid lines denote  mean values.
      Markers show aggregation periods ($15$\,\unit{min}, $1$\,\unit{hr}, $12$\,\unit{hr}, $1$\,\unit{d}, $5$\,\unit{d}, $10$\,\unit{d}, and $15$\,\unit{d}).
       }
    \label{fig:multi_run}
  \end{figure}

The second numerical experiment aims to evaluate the role of channel shape and friction factors on the findings above and to examine the effect of variability in water discharge.
Across the range of channel shapes and friction factors  examined (Section\,\ref{sect:imp}), variability in water velocity and shear stress remains higher in the subglacial system compared to the subaerial one in both \alpine{} and \icesheet{}.
This suggests that there is likely greater intermittency in sediment transport in subglacial systems compared to subaerial ones, especially in supply-limited systems where sediment could be exhausted or mobilized over diurnal discharge cycles.
In contrast, in \alpine{}, the variability in width-integrated shear stress is comparable between the subaerial and subglacial systems, due to the highly variable water discharge, largely due to diurnal variations (Figure~\ref{fig:multi_run}).
However, in \icesheet{}, with lower water discharge fluctuations, the variability in width-integrated shear stress remains consistently higher in the subglacial system across the time aggregation periods.

Results from  both \alpine{} and \icesheet{} scenarios show that the variability in subglacial  decreases with aggregations longer than approximately 1--5 days (Figure~\ref{fig:multi_run}).
Yet, substantial differences in the variability between the model outputs for subglacial and subaerial cases persist in periods of up to $15$ days, after which seasonal variations might be conflated.
Note that the timing of peak sediment transport conditions still occurs across these periods, and they are are thus not captured in aggregating water discharge over longer periods (Figure~\ref{fig:model_outs}).

\section{Discussion}

\subsection{Hydrology and evolving channel shape in driving sediment transport variability}
\label{sect:dis_qsc}

@@

Sediment transport capacity in subglacial channels responds more rapidly to changes in water discharge compared to their subaerial counterparts \cite{alley1997}.
In subglacial R\"othlisberger channels in steady state with water discharge, the width-integrated shear stress response to water discharge is $ w_g\,\tau_{g}\, \propto\,  Q^{\sim\frac{4}{5}}$ (Table\,\ref{tab:eqs1}).
However, subglacial channels rarely operate in a steady state \cite<e.g.>{gimbert2016}.
For sufficiently short periods, channel width and wetted area are fixed in the subglacial system (akin to pipe flow) so that the response of width-integrated shear stress is $w_g\, \tau_{g}\, \propto\,  Q^{2}$ (Table\,\ref{tab:eqs1}).
\ref{sect:scaling} provides a complete description of sediment transport capacity in different drainage conditions.
Note that in pipe flow conditions, maximum sediment transport would coincide with peak water discharge and a co-varying relationship with water discharge would occur, unlike with the evolving channel size here (Section~\ref{sect:sub_mode}; Figure \ref{fig:model_outs} and \ref{fig:Qw_vari}). 
Similar adjustments in channel shape occur in subaerial channels \cite{phillips2016}, albeit at timescales longer than hours to days in subglacial channels.

Substantial differences emerge in model outputs between the different hydrological forcings of the \alpine{} and \icesheet{} cases.
Indeed, the more independent relationship between model outputs and water discharge in \alpine{} results from the
relatively larger discharge variations on daily to weekly timescales compared to \icesheet{}.
This occurs as the evolution of the R-channel's size is small compared to the variations in water discharge.
Such water discharge variability in alpine catchments may cause shear stress to approach the water discharge squared \cite< \ref{sect:scaling}; c.f.>{alley1997}.
With respect to \icesheet{}, a closer approximation of the hydraulics could be represented by pipe flow here, where the channel size remains fixed.
Conversely, the reduced variability in water discharge in \icesheet{} may result in the R-channel's size being closer to equilibrium concerning the variations in water discharge.
In this case, the exponent on water discharge for shear stress is likely less than $2$, but greater than $\frac{4}{5}$ occurring when the R-channel's size evolves in equilibrium with water discharge (see \ref{sect:scaling}).
As a result, a stronger correlation persists between subglacial model outputs and water discharge in \icesheet{} compared to \alpine{} (Figure~\ref{fig:Qw_vari}).

@@

The stronger link between sediment transport capacity and water discharge in subaerial channels may mean that variations in sediment transport records can be more easily attributed to hydro-climatic conditions than in subglacial systems.
A glacier's sediment transport capacity is impacted by the ice thickness controlling the channel closure rate, and the glacier's surface slope, in addition to water discharge and sediment size \cite<Figure~\ref{fig:multi_run}, Section~\ref{sect:sub_mode}; >{rothlisberger1972,shreve1972,gimbert2016,stevens2022,walder1994}.
In a supply-limited state, then glaciers' sediment discharge record also responds to  sediment availability and bedrock erosion from  water pressure variations and sliding  \cite<e.g.>{iverson2012,herman2015,delaney2023}.
This multitude of processes lies in contrast to many subaerial channels, where transport capacity typically responds to sediment size, channel shape, water discharge, and hydraulic gradient.
The hydraulic gradient and channel shape can remain stable over years to millennia \cite<Section~\ref{sect:fluv}; e.g.>{tucker1997,wickert2019}.

@@

Furthermore, the greater variations in subglacial sediment transport capacity in both catchment types compared to subaerial ones could cause a supply-limited regime to persist at many glaciers.
In subglacial systems, sediment's threshold of motion can be reached more frequently and across a range of water discharges, compared to subaerial systems (Figure~\ref{fig:Qw_vari}).
Sediment exhaustion through this process may  explain the stronger dependence of sediment discharge from the Greenland Ice Sheet on basal shear stress, a proxy for bedrock erosion, rather than glacier melt \cite{overeem2017}.

\subsection{Parameterizing  sediment transport in glacierized catchments}

Numerical models of subglacial sediment transport generally apply classical parameterizations of subglacial hydrology to evaluating sediment transport conditions \cite<e.g>{hewitt2019,beaud2018}.
In these R-channel models, viable water pressure regimes can be achieved by changing the shape and friction factor of the channels, enabling assumptions of semi-circular channels in most applications \cite{werder2010b}.
However, the role of channel width in the findings here suggests that more thorough parameterization of channel shape and channel width is needed in assessing subglacial sediment transport capacity as compared to water pressure (Section\,\ref{sect:sub_mode}, \ref{sect:scaling}).

In applying these models to different climate scenarios, water discharge measurements at the sub-hour timescale remain unavailable in many catchments.
The model outputs' variability in subglacial channels approaches that of subaerial ones  over $1-5$ days (Figure~\ref{fig:multi_run}), possibly resulting in parity between subaerial and subglacial channels over these periods.
This suggests that aggregating water discharge over daily to weekly periods could yield similar variations between subglacial and subaerial systems, although generally steeper hydraulic gradients of glaciers may cause elevated sediment transport capacity there (Figure\,\ref{fig:model_outs}). 
Yet, water discharge over periods longer than days may omit key processes, such as the time of peaks in sediment transport capacity (Figure~\ref{fig:model_outs}).

@@

The sediment transport capacity of a channel is driven by its width and the shear stress exerted by the flowing water, which is proportional to the square of the flow velocity.
Subaerial channels alter both their channel width and water velocity immediately in response to changing water discharge.
In contrast, pressurized subglacial channels largely accommodate rapidly changing water discharge by altering water velocity.
Their size responds over several days to changes in discharge.
Thus the response of sediment transport capacity is much more sensitive to changes in water discharge in subglacial channels compared to subaerial ones.

Implications of the presented findings include:
\begin{enumerate}
\item the changing relationship between water discharge and sediment transport capacity below glaciers creates larger variability in sediment transport capacity compared to river systems;
  \vspace{.1cm}

\item  water discharge variability in the two hydrographs impacts subglacial sediment transport capacity variations;
  \vspace{.1cm}

\item the uneven response of sediment transport capacity to water discharge variations shows a different response and timing to periods of elevated water discharge in subglacial channels compared to subaerial ones.
  However, representing subglacial sediment transport capacity with either open-channel flow or pipe flow neglects variations in timing.
\end{enumerate}
